Biden More Favorable Among Female, Black, Asian, Hispanic, Middle Eastern, and Non-gun Owning Voters in the 2020 US Presidential Election

### 2024-03-17 paper/paper.qmd
Fix grammatical errors

The United States (US) presidential election of 2020 resulted in President Joe Biden’s victory, making him the 46th President of the US. The two top runners were Biden for the Democrats and former President Donald Trump for the Republicans. The presidential race between the two was of significant interest to statisticians and polling experts due to its potential impact on polling models and election forecasting. To provide some background, President Biden ran on his merit as a former Vice President of eight years and had faced lots of criticism during his time in office. Moreover, Former President Trump had controversial and career-damaging events. Before the 2020 elections, Trump was charged with misconduct while in office but the senate acquitted him of those misconduct charges [@impeachment]. Thus, both Biden and Trump faced have previously criticism from the electorate. An interesting question to consider is despite these, what characteristics did the voters have? Do they tell us a story on the type of people that voted for Biden and the type that voted for Trump? Answering these questions can help develop polling models and future election forecasting.

For our paper, we used a logistic regression model to estimate the likelihood of individuals voting for Biden or Trump, given certain characteristics. We use a binary outcome variable and three predictor variables. Our estimand is the probability that an individual voted for Biden or Trump based on three characteristics; their gender, race, and ownership of a gun. The reason for this choice will be further examined in @sec-model. This data driven approach to elections can bring insight to voter characteristics and the importance of survey data and knowing the electorate. We sought to produce results that could be used to predict voter behavior in the future.

Based on our model and analysis, we found that certain demographic groups were more include to support either candidate. Our findings revealed that women were more likely to vote for Biden compared to Trump. This finding is consistent with broader trends in political behavior, as gender has previously shown to influence voting preferences. Biden’s democratic policies were more convincing to women and will be discussed in @sec-discussion. Another major finding was that Black, Asian, Hispanic, and Middle Eastern individuals are also more likely to vote for Biden. Biden has been widely known to run on a more progressive platform, giving importance on racial justice and addressing racial disparities. Our finding emphasizes that people from diverse and ethnic backgrounds are more inclined to vote for progressive representatives. Our final major finding was that non-gun owners are also more likely to vote for Biden. Gun owners often have strong opinions on gun control policies. Our finding shows that those who do not possess guns favored Biden’s gun control and other firearm-related policies, as opposed to Trump’s policies.

@@

From the dataset, we selected `gender`, `race`, `gunown` from the pre-election questionnaire. In @fig-race-gender, we can see support for candidate based on gender and race. For gender, we were given two options in the study, male or female. For race, we considered eight options, White, Black, Hispanic, Asian, Native American, Middle Eastern, two or more races, and other. Note that only one selection was allowed in the original study for both gender and race. @fig-gun-ownership-gender illustrates the support for each candidate based on gun ownership status, with four options. Individuals who personally own a gun, who don’t personally own one, but someone in their household owns a gun, no one in the household owns a gun, and not sure. We also selected a variable from the post-election questionnaire, the variable that recorded whether an individual voted for Biden or Trump in the 2020 Presidential election, `CC20_410`. This allowed us to establish a binary outcome variable for the probability of voting Biden or Trump in our model in @sec-model.

>>>>

The United States (US) presidential election of 2020 resulted in President Joe Biden’s victory, making him the 46th President of the US. The two top runners were Biden for the Democrats and former President Donald Trump for the Republicans. The presidential race between the two was of significant interest to statisticians and polling experts due to its potential impact on polling models and election forecasting. To provide some background, as former Vice President, President Biden had faced lots of scrutiny during his time in office. Likewise, former President Trump had controversial and career-damaging events. Before the 2020 elections, Trump was charged with misconduct while in office, but was acquitted by the Senate [@impeachment]. Thus, both Biden and Trump have previously faced criticism from the electorate. An interesting question to consider is what characteristics did the voters have? Do they tell us a story on the type of people that voted for Biden and the type that voted for Trump? Answering these questions can help develop polling models and future election forecasting.

For our paper, we used a logistic regression model to estimate the likelihood of individuals voting for Biden or Trump, given certain characteristics. We used a binary outcome variable and three predictor variables. Our estimand is the probability that an individual voted for Biden or Trump based on three characteristics; their gender, race, and ownership of a gun. The reason for this choice will be further examined in @sec-model. This approach to elections can bring insight to voter characteristics and the importance of survey data and knowing the electorate. We sought to produce results that could be used to predict voter behavior in the future.

Based on our model and analysis, our findings revealed that women were more likely to vote for Biden compared to Trump. This finding is consistent with broader trends in political behavior, as gender has previously shown to influence voting preferences. Biden’s democratic policies were more convincing to women and will be discussed in @sec-discussion. Another major finding was that Black, Asian, Hispanic, and Middle Eastern individuals were also more likely to vote for Biden. Biden has been widely known to run on a more progressive platform, giving importance on racial justice and addressing racial disparities. Our finding emphasizes that people from diverse and ethnic backgrounds are more inclined to vote for progressive representatives. Our final major finding was that non-gun owners were also more likely to vote for Biden. Gun owners often have strong opinions on gun control policies. Our finding shows that those who do not possess guns favored Biden’s gun control and other firearm-related policies, as opposed to Trump’s policies.

@@

From the dataset, we selected `gender`, `race`, and `gunown` from the pre-election questionnaire. In @fig-race-gender, we can see support for the candidate based on gender and race. For gender, we were given two options in the study, male or female. For race, we considered eight options: White, Black, Hispanic, Asian, Native American, Middle Eastern, two or more races, and other. Note that only one selection was allowed in the original study for both gender and race. @fig-gun-ownership-gender illustrates the support for each candidate based on gun ownership status, with four options. Individuals who personally own a gun, who don’t personally own one, but someone in their household owns a gun, no one in the household owns a gun, and not sure. We also selected a variable from the post-election questionnaire, the variable that recorded whether an individual voted for Biden or Trump in the 2020 Presidential election, `CC20_410`. This allowed us to establish a binary outcome variable for the probability of voting Biden or Trump in our model in @sec-model.
